Pipit on the Post: Proving Pre- and Post-conditions of Reactive Systems

### 2024-03-28 2024pipit/section/60-Evaluation.tex
copy and tighten

To evaluate Pipit, we have implemented the high-level logic of a time-triggered Controller Area Network (CAN) bus driver~\cite{ISO11898_4}.
The CAN bus is commonly found in safety-critical automotive and industrial settings.
The time-triggered network architecture defines a static schedule of network traffic.
All nodes on the network must adhere to the same schedule, which significantly increases the reliability of periodic messages~\cite{fuehrer2001time}.

At a high level, the schedule is described by a system matrix which consists of rows of basic cycles.
Each basic cycle consists of a sequence of actions to be performed at particular time-marks.
Actions in the schedule may not be relevant to all nodes, so each node has its own local array containing the relevant triggers; trigger actions include sending and receiving application-specific messages, sending reference messages, and triggering `watch' alerts.
The trigger action for receiving an application-specific message checks that a particular message has been received since the trigger was last executed; depending on this, the driver increments or decrements a message-status-counter, which will in turn signal an error once the upper limit is reached.
Reference messages start a new basic cycle and are used to synchronise the nodes.
Watch alerts are generally placed after the expected end of the cycle and are used to signal an error if no reference message is received.

The TTCAN protocol can be implemented in two levels of increasing complexity.
In the first level, reference messages contain the index of the newly-started cycle.
In the second level, the reference messages also contain the value of a global fractional clock and whether any gaps have occurred in the global clock, which allows other nodes to calibrate their own clocks.
We implement the first level as it is more amenable to software implementation~\cite{hartwich2002integration}.

@@

We compare against two Kind2 implementations: one corresponds closely to the Pipit development, while the other includes a critical simplification to modify the trigger-enabled set to be a single cycle index.
In the specification, the enabled set is implemented as a cycle-offset and repeat-factor.
Checking if a trigger is enabled in the current cycle requires nonlinear arithmetic, which is difficult for SMT solvers.
In our Pipit development, we can treat the definition of the cycle set abstractly.
However, in the Kind2 development, quantified formulas cannot contain function calls, which means that we cannot hide the implementation of the enabled-set check by providing an abstract contract.
This limitation also makes the specification quite unwieldy, as functions must be manually inlined.

\autoref{f:evaluation:kind2-runtime} shows the verification runtime for different sizes of arrays; the Pipit version is parametric in the array size, and is thus verified for all sizes of arrays.
We ran these experiments in Docker on an Intel i5-12500 with 32GB of RAM.
Both Kind2 and Pipit developments of the full trigger-fetch logic are roughly the same size, on the order of two-hundred lines of code including comments.
Ignoring whitespace and comments, the Pipit implementation has 26 lines of actual executable code, while the Kind2 code has 32.
The majority of the remaining code comprises the definition of valid schedules (34 for Pipit, 28 for Kind2), and the lemma statements and invariants (12 for Pipit, 31 for Kind2), as well as contract statements and boilerplate.

@@

\begin{figure}
  \center
\begin{tabular}{r|rr|rr|rr}
  & \multicolumn{4}{c|}{Kind2} & Pipit \\
  & \multicolumn{2}{c|}{simple enable-set} & \multicolumn{2}{c|}{full enable-set} & \\
  size & wall-clock & CPU-time & wall-clock & CPU-time & wall-clock & CPU-time \\
  \hline

@@

\end{tabular}
\caption{Verification time for trigger-fetch; simple enable-set uses a simplified version of the enable-set, while full enable-set uses bitwise arithmetic as in the TTCAN specification.
The ``wall-clock time'' denotes the elapsed time that an engineer must spend waiting for the result; the CPU time denotes the total time spent computing by all of the CPU cores.
The verification time for Pipit is a once-and-for-all proof that is parametric in the size of the array.
The time limit was one hour.}
\label{f:evaluation:kind2-runtime}
\end{figure}

>>>>

To evaluate Pipit, we have implemented the high-level logic of a Time-Triggered Controller Area Network (TTCAN) bus driver~\cite{ISO11898_4}, described earlier in \autoref{s:motivation}.
The CAN bus is common in safety-critical automotive and industrial settings.
The time-triggered network architecture defines a static schedule of network traffic;
by having all nodes on the network adhere to the schedule, the reliability of periodic messages is significantly increased~\cite{fuehrer2001time}.

The TTCAN protocol can be implemented in two levels of increasing complexity.
In the first level, reference messages, which perform synchronisation between nodes, contain the index of the newly-started cycle.
In the second level, the reference messages also contain the value of a global fractional clock and whether any gaps have occurred in the global clock, which allows other nodes to calibrate their own clocks.
We implement the first level as it is more amenable to software implementation~\cite{hartwich2002integration}.

@@

We compare against two Kind2 implementations: one corresponds closely to the Pipit development, while the other includes a critical simplification to modify the trigger-enabled set to be a single cycle index.
In TTCAN proper, the enabled set is implemented as a cycle-offset and repeat-factor.
Checking if a trigger is enabled in the current cycle requires nonlinear arithmetic, which is difficult for SMT solvers.
In our Pipit development, we can treat the definition of the cycle set abstractly.
However, in the Kind2 development, quantified formulas cannot contain function calls, which means that we cannot hide the implementation of the enabled-set check by providing an abstract contract.
This limitation also makes the specification quite unwieldy, as we must manually inline any functions in quantified formulas.

\autoref{f:evaluation:kind2-runtime} shows the verification runtime for different sizes of arrays; the Pipit version is parametric in the array size, and is thus verified for all sizes of arrays.
We ran these experiments in Docker on an Intel i5-12500 with 32GB of RAM.
Both Kind2 and Pipit developments of the full trigger-fetch logic are roughly the same size, on the order of two-hundred lines of code including comments.
Ignoring whitespace and comments, the Pipit implementation of trigger-fetch has 26 lines of actual executable code, while the Kind2 code has 32.
The majority of the remaining code comprises the definition of valid schedules (34 for Pipit, 28 for Kind2), and the lemma statements and invariants (12 for Pipit, 31 for Kind2), as well as contract statements and boilerplate.

@@

\begin{figure}
  \center
\begin{tabular}{r|rr|rr|rr}
  & \multicolumn{4}{c|}{Kind2} & Pipit \\
  & \multicolumn{2}{c|}{simple enable-set} & \multicolumn{2}{c|}{full enable-set} & \\
  size & wall-clock & CPU time & wall-clock & CPU time & wall-clock & CPU time \\
  \hline

@@

\end{tabular}
\caption{Verification time for trigger-fetch; simple enable-set uses a simplified version of the enable-set, while full enable-set uses bitwise arithmetic as in the TTCAN specification.
The wall-clock time denotes the elapsed time that an engineer must spend waiting for the result; the CPU time denotes the total time spent computing by all of the CPU cores.
The verification time for Pipit is a once-and-for-all proof that is parametric in the size of the array.
The time limit was one hour.}
\label{f:evaluation:kind2-runtime}
\end{figure}

### 2024-03-28 2024pipit/section/70-Related.tex
copy and tighten

Verifying the time-triggered CAN driver is trickier, as the restrictions placed on the triggers array --- that triggers are sorted by time-mark, there must be an adequate time-gap between a trigger and its next-enabled, and a trigger's time-mark must be greater-than-or-equal-to its index --- naturally require quantifiers.
As described in \autoref{s:evaluation}, the Kind2 does include experimental array and quantifier support, but in our experiments was limited to verifying small arrays up to 8 or 16 triggers.
Additionally, due to the limitations on top-level array definitions, compiling the program with Lustre~V6 would result in the entire triggers array being copied to the stack each iteration.

Other model-checkers for Lustre such as Lesar~\cite{raymond2008synchronous}, JKind~\cite{gacek2018jkind} and the original Kind~\cite{hagen2008scaling} do not support quantifiers.
It may be possible to encode the quantifiers as fixed-size loops, but ensuring that these loops do not affect the execution or runtime complexity of the generated code does not appear to be straightforward.

>>>>

Verifying the time-triggered CAN driver is trickier, as the restrictions placed on the triggers array --- that triggers are sorted by time-mark, there must be an adequate time-gap between a trigger and its next-enabled, and a trigger's time-mark must be greater-than-or-equal-to its index --- naturally require quantifiers.
As described in \autoref{s:evaluation}, Kind2 does include experimental array and quantifier support, but in our experiments was unable to verify the full logic for arrays up to the 64 triggers, which is the size supported by hardware implementations of TTCAN.
Additionally, due to the limitations that require the constant triggers array to be passed as an argument, compiling the program with Lustre~V6 would result in the entire triggers array being copied to the stack each iteration.

Other model-checkers for Lustre such as Lesar~\cite{raymond2008synchronous}, JKind~\cite{gacek2018jkind} and the original Kind~\cite{hagen2008scaling} do not support quantifiers.
It may be possible to encode the quantifiers as fixed-size loops in those that support arrays, but ensuring that these loops do not affect the execution or runtime complexity of the generated code does not appear to be straightforward.

### 2024-03-28 2024pipit/section/80-Conclusion.tex
copy and tighten

We have presented Pipit, a verified compiler and proof system for reactive systems.
Our implementation of the TTCAN driver logic shows that, by embedding pure \fstar{} functions for array operations, Pipit can express programs which are currently unsupported by other verified Lustre compilers.
Pipit can also verify high-level program properties which are difficult to express and prove in existing Lustre model-checkers.
Our development includes verified translations to both abstract and executable transition systems; both are shown to preserve the dynamic semantics.
We also introduced a checked semantics, which describes the proof obligations of a program; in future work, we intend to verify that the proof obligations generated by the abstract transition system match the checked semantics.

>>>>

We have presented Pipit, a verified compiler and proof system for reactive systems.
Our implementation of the TTCAN driver logic shows that, by embedding pure \fstar{} functions for array operations, Pipit can express programs which are currently unsupported by other verified Lustre compilers.
Pipit can also verify high-level program properties which are difficult to express and prove in existing Lustre model-checkers.
Our development includes verified translations to both abstract and executable transition systems; both are shown to preserve the dynamic semantics.
We also introduced a checked semantics, which describes the semantics of checked properties and contracts; proof obligations generated by translation to abstract transition system are verified to correspond to these semantics.
